MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs

### 2025-07-03 latex/paper/sections/10-placement.tex
more reorg

In more sophisticated placers, the proposed movement can also be selected by finding the centroid between \texttt{SiteInst}s that share a net with the current object, or even a hybrid of random and centroid selection. 
Moves that actively attempt to predict a better placement are often referred to as "directed" moves in literature. 
Placers that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" placers and are akin to physics simulators.

In force directed placers, there is a concept of global placement followed by detailed placement
Global placement places objects freely on an analog plane and is driven by directed movement or analytical solution.
In global placement, objects do not have to adhere to any hardware constraints and can even overlap with one another.
Detailed placement takes the global placement and attempts to fit it within the constraints of the hardware. 
Detailed placement "legalize" the global placement by discretizing the analog positions of the objects onto the discrete grid of resources of the \texttt{device}.
Overlapping objects are spread out until every \texttt{SiteInst}-\texttt{Cell} is assigned its own unique \texttt{Site}-\texttt{BEL}.
The process of spreading objects somewhat resembles primitive fluid dynamics simulators or thermal diffusion simulators, where packets of energy are spread around an object until each voxel or pixel meets a maximum threshold of energy.

There also exists other strategies to mix packing, placement, and legalization in different ways. (CITE THAT ONE PAPER).
In this context, our placer follows a straightforward pack-legalize-place strategy.

@@

\newcolumn
\begin{lstlisting}[language=java, caption={SA pseudocode: outer loop}, label={lst:sa_outer}]
public void placeDesign(PackedDesign packedDesign) {
    randomInitialPlacement(packedDesign);
    // init_temp=10000, alpha=0.98, max_passes=300
    initCoolingSchedule(10000, 0.98, 300)
    int passes = 0;
    while (passes < 300) {
        updateTemperature(passes);
        moveAll(packedDesign)
        passes++;
    }
}

@@

{
    \centering
    \includegraphics[width=\columnwidth]{figures/placement/swapSingleSite.png}
    \captionof{figure}{Single Site Swap Proposal}
    \label{fig:swapSingleSite}
}
\vfill

\begin{lstlisting}[language=java, caption={Cooling Schedule and Move Acceptance with Temperature}, label={lst:sa_acceptance}]
protected List<Double> coolingSchedule;
public void initCoolingSchedule(double initialTemp, double alpha, int movesLimit) {
    double currentTemp = initialTemp;
    for (int i = 0; i < movesLimit; i++) {
        this.coolingSchedule.add(currentTemp);
        currentTemp *= alpha; // geometric cooling
    }
}

@@

>>>>

@@

\begin{lstlisting}[language=java, caption={SA pseudocode: outer loop}, label={lst:sa_outer}]
public void placeDesign(PackedDesign packedDesign) {
    randomInitialPlacement(packedDesign);
    // init_temp=10000, alpha=0.98, max_passes=300
    initCoolingSchedule(10000, 0.98, 300)
    int passes = 0;
    while (passes < 300) {
        updateTemperature(passes);
        moveAll(packedDesign)
        passes++;
    }
}

@@

{
    \centering
    \includegraphics[width=0.9\columnwidth]{figures/placement/swapSingleSite.png}
    \captionof{figure}{Single Site Swap Proposal}
    \label{fig:swapSingleSite}
}

\begin{lstlisting}[language=java, caption={Cooling Schedule and Move Acceptance}, label={lst:sa_acceptance}]
protected List<Double> coolingSchedule;
public void initCoolingSchedule(double initialTemp, double alpha, int movesLimit) {
    double currentTemp = initialTemp;
    for (int i = 0; i < movesLimit; i++) {
        this.coolingSchedule.add(currentTemp);
        currentTemp *= alpha; // geometric cooling
    }
}

@@

\subsection{Variations}
\label{subsec:variations}

In more sophisticated placers, the proposed movement can also be selected by finding the centroid between \texttt{SiteInst}s that share a net with the current object, or even a hybrid of random and centroid selection. 
We will implement variations on the vanilla SA algorithm that use centroid selection. 
Moves that actively attempt to predict a better placement are often referred to as "directed" moves in literature. 
Placers that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" placers and are akin to physics simulators.

In force directed placers, there is a concept of global placement followed by detailed placement.
Global placement places objects freely on an analog plane and is driven purely by directed movement and analytical solution.
In global placement stage, only non-hardware related objectives are considered. 
Objects do not have to adhere to any hardware constraints and can even overlap with one another.
Afterwards, detailed placement "legalizes" the global placement by discretizing the analog positions of the objects onto the discrete grid of resources of the \texttt{device}.
Overlapping objects are iteratively spread out until every \texttt{SiteInst}-\texttt{Cell} is assigned its own unique \texttt{Site}-\texttt{BEL}.
The spreading process somewhat resembles primitive fluid dynamics simulators or thermal diffusion simulators, where packets of energy are spread through a medium until each voxel or pixel of the medium falls below a maximum threshold of energy. 
There also exists other strategies to mix packing, placement, and legalization in different orders. 
In this context, our placer follows a very straightforward pack-legalize-place strategy with SA.

### 2025-07-03 latex/paper/sections/11-placement_results.tex
more reorg

We will need an HDL design that utilizes a good mixture of LUTs, FFs, BRAMs, and DSPs at scale to demonstrate the robustness and performance of our placers. 
Any DSP subsystem can serve as a good candidate for such a demonstration. 
Here, we will perform placement on a 2048th-order FIR Filter which was conveniently created as coursework for a VLSI course. 
After passing synthesis, the FIR Filter design calls for the primitive cells shown in the second column of table \ref{fig:utilization}.
We will perform placement of this design on a \texttt{xc7z020} FPGA, which is a mid-ranged Xilinx device containing the available BELs shown in the first column of table \ref{fig:utilization}.
Note that \texttt{LUT5-LUT6} pairs are counted as a single \texttt{LUT}.
% Also note that the number of \texttt{CARRY4} BELs is equal to the total number of \texttt{SLICEL} and \texttt{SLICEM} Sites combined and that the number of \texttt{LUT}s is equal to four times that amount while the number of \texttt{FF}s is equal to eight times that amount.

@@

\begin{itemize}
    \item \texttt{PlacerGreedyRandom}: all undirected random moves with greedy acceptance (always zero temperature)
    \item \texttt{PlacerGreedyMidpoint}: all directed centroid moves with greedy acceptance (always zero temperature)
    \item \texttt{PlacerAnnealRandom}: all undirected random moves with annealing acceptance \textbf{(the basic SA algorithm)}
    \item \texttt{PlacerAnnealMidpoint}: all directed centroid moves with annealing acceptance
    \item \texttt{PlacerAnnealHybrid}: \textbf{Initial:} 50\% random - 50\% centroid moves with annealing acceptance. \textbf{At Near-Zero Temperature:} 100\% centroid moves with annealing acceptance.
\end{itemize}

((Talk about the initial cost.))

@@

The next three placers performed significantly better than the previous two placers.
Our vanilla SA placer, \texttt{PlacerAnnealRandom} is represented by the green curve.
We can observe that the cost decreases more steadily than the previous two placers, with noticeable spikes in total system cost due to hill-climbing, until settling at the 346K mark as the global temperature cools to zero.

@@

{
    \centering
    \includegraphics[width=\columnwidth]{figures/results/combined_cost_history_linear.png}
    \captionof{figure}{All placers overlaid}
    \label{fig:placers_overlay}
}

The best performing placer was \texttt{PlacerAnnealHybrid}, represented by the blue curve.
We can observe that \texttt{AnnealHybrid} settles at a similar rate with the vanilla \texttt{AnnealRandom}, but eventually settles at a lower minimum at 322K, outperforming even \texttt{GreedyRandom}.

@@

\newcolumn
{
    \centering
    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealRandom/random_placement.png}
    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealRandom/00000010.png}
    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealRandom/00000100.png}
    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealRandom/00000299.png}
    \captionof{figure}{\texttt{PlacerAnnealRandom}}
    \label{fig:PARSnapshots}
}

>>>>

We will need an HDL design that utilizes a good mixture of LUTs, FFs, BRAMs, and DSPs at scale to demonstrate the robustness and performance of our placers. 
Any DSP subsystem can serve as a good candidate for such a demonstration. 
Here, we will perform placement on a 2048-order FIR Filter which was conveniently created as coursework for a VLSI course. 
After passing synthesis, the FIR Filter design calls for the primitive cells shown in the second column of table \ref{fig:utilization}.
We will perform placement of this design on a \texttt{xc7z020} FPGA, which is a mid-ranged Xilinx device containing the available BELs shown in the first column of table \ref{fig:utilization}.
Note that \texttt{LUT5-LUT6} pairs are counted as a single \texttt{LUT}.
% Also note that the number of \texttt{CARRY4} BELs is equal to the total number of \texttt{SLICEL} and \texttt{SLICEM} Sites combined and that the number of \texttt{LUT}s is equal to four times that amount while the number of \texttt{FF}s is equal to eight times that amount.

@@

\begin{itemize}
    \item \texttt{PlacerGreedyRandom}: all undirected random moves with greedy acceptance
    \item \texttt{PlacerGreedyMidpoint}: all directed centroid moves with greedy acceptance
    \item \texttt{PlacerAnnealRandom}: all undirected random moves with annealing acceptance \textbf{(the basic SA algorithm)}
    \item \texttt{PlacerAnnealMidpoint}: all directed centroid moves with annealing acceptance
    \item \texttt{PlacerAnnealHybrid}: \textbf{Initial:} 50\% random - 50\% centroid moves with annealing acceptance. \textbf{At Near-Zero Temperature:} 100\% centroid moves with annealing acceptance.
\end{itemize}

@@

The next three placers performed better than the previous two placers by decent margin.
The vanilla SA placer, \texttt{PlacerAnnealRandom}, is represented by the green curve.
We can observe that the cost decreases more steadily than the previous two placers, with noticeable spikes in total system cost due to hill-climbing, until settling at the 346K mark as the global temperature cools to zero.

@@

{
    \centering
    \includegraphics[width=0.9\columnwidth]{figures/results/combined_cost_history_linear.png}
    \captionof{figure}{All placers overlaid}
    \label{fig:placers_overlay}
}

The best performing placer was \texttt{PlacerAnnealHybrid}, represented by the blue curve.
Starting at zero passes, the placer randomly selects between random selection and centroid selection, allowing the placer to make on average more cost-reducing moves, while still maintaining some diversity of moves for hill-climbing. 
Then, at near-zero temperature, the placer makes exclusively centroid moves with annealing acceptance to speed up the final crystallization into the local minimum. 
With centroid moves and annealing acceptance, hill-climbing can only be achieved via swaps.
We can observe that \texttt{AnnealHybrid} settles at a similar rate with the vanilla \texttt{AnnealRandom}, but eventually settles at a lower minimum at 322K, slightly outperforming \texttt{GreedyRandom}.

@@

{
    \centering
    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealRandom/random_placement.png}
    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealRandom/00000010.png}
    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealRandom/00000100.png}
    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealRandom/00000299.png}
    \captionof{figure}{\texttt{PlacerAnnealRandom}}
    \label{fig:PARSnapshots}
}
